---
title: "Examples"
author: Maxime Jan
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    fig_width: 6
    fig_height: 4
    dev: jpeg
---

# Sleep-Wake Driven Models, an R package [SWDMr]

*An R package to fit models for sleep-wake driven phenotypes* 

- [x] Read and format vigilant state data
- [x] Fit a Process-S dynamic model
- [x] Fit an driven damped harmonic oscillator model
- [x] Add solution Anne Skeldon


Statistical methods are inspired by Jake Yeung method for model selection. See [J.Yeung github](https://github.com/jakeyeung/SleepDepAnalysis) and [PNAS publication](https://doi.org/10.1073/pnas.1910590116)


```{r}
library(SWDMr) # Package for model construction, objective function building

# External package
library(optimx) # Package for parameter optimization
library(ggplot2) # Package for visualization
library(patchwork)
```


# Format vigilant state

Get vigilant state of mice. Each row is a 4 seconds epoch containing the vigilant state of the mouse

- Wake = w | 1
- NREM sleep = n | 2
- REM sleep = r | 3

```{r}
data(SleepWakeData)
head(as.data.frame(SleepWakeData))
```

Compute the amount of wake,nrem and rem sleep per 6 minutes bin (0.1h)

```{r}
# Read the matrix and transform it into a data.frame for time spent in each state per "concattimesec" bin [in sec]
SWdf<-Read_SW(SleepWakeData,concattimesec = 360,ResultBy ="proportion") # 300 = 5min, 180 = 3 min, 360 = 6min

# We can add day and night cycle
SWdf<-SWdf_AddLD(SWdf)

# We merge day 1 and 2 that are "baseline", then following days are 3 and 4, then baseline sleep is reiterated
SWdf<-SWdf_DayMerging(SWdf,Daysformat=list(c(1,2),c(1,2),3,4,c(1,2)),concattimesec=360)

# We add sleep deprivation even at time 48
SWdf<-SWdf_AddSD(SWdf,c(48,54))

# End result:
head(SWdf)
```

Sleep amount in the experiment

```{r}
gg<-ggplot(aes(x=Time,y=Sleep),data=SWdf)+geom_line(color="darkblue",size=1)
gg<-gg+scale_x_continuous(breaks=seq(0,96,by=12))
gg<-gg+ylab("Sleep amount [h]")+xlab("Time [h]")+ theme_bw()
gg
```



# Fit a Driven Damped harmonic oscillator model

In this model, sleep and wake change the phenotype speed and act as a force on the oscillator.

We also use a second force driving the oscillator in the form of a sine-wave representing the
force applied by the circadian systems to synchronize phenotypes to the current time.

In this example, we model the dynamics of Arntl (Bmal1)

## Data

First we load data for time spent awake or asleep


```{r}
# Contain a data.frame of time spent asleep or awake
data(SWdf)
head(SWdf)
```


We then load phenotypes containing also time of sampling corresponding with the sleep-wake data.frame

```{r}
data(RNAExpression)
head(RNAExpression)
```


## Create model

We create a swdmr object containing sleep-wake data and phenotypes

```{r}
swdmr <- SWDMr(SWdist=SWdf, Gexp=RNAExpression)
swdmr
```

Simulation goes up to T96 but some points in RNA expression data.frame goes up to T222


We then initiate a model for a Driven Damped Harmonic Oscillator (DDHO)

```{r,warning=F}
model<-initDDHOmodel(swdmr,VarExp = "Arntl")
```

In our model, we fix the equilibrium position to the mean expression of Arntl in baseline

```{r}
MeanGeneExprInBaseline<-mean(RNAExpression$Arntl[RNAExpression$Time<=48])
model<-FixIntercept(model,MeanGeneExprInBaseline)
```

The force driving our oscillator are Sleep-Wake

```{r}
model<-AddForce(model,"Wake")
model<-AddForce(model,"Sleep")
```

A second force is used, representing the sychronizing force with biological time (CT). 
As mouse are under 12:12 LD cycle, this force has a 24h period

```{r}
model<-AddSinF(model,FixPer = 24)
```


Our model start at the equilibrium position with a speed of 0

```{r}
model<-SetYinitMode(model,mode = "Intercept_0")
```

We replicate baseline sleep-wake distribution to allow the oscillator to reach a stable oscillation

```{r,warning=F}
model<-ReplicateDrivingForce(model,c(0,24.0),40)
```


```{r}
model<-SetParametersModel(model)
```


This is our model:

```{r}
model
```


## Optimization

We create an objective function:

```{r}
objfun<-SWDMrGetEvalFun(model)
```

We use optimx to optimize the free parameter to minimize the RSS

```{r}
# We assume the following starting parameters:

# Natural frequency of our oscillator close to 24h
# low damping
params<-c(omega=2*pi/24,loggamma=-1,Wake=0,Sleep=0,AmpSin=0,PhiSin=pi)
# Many other algorithms and options exist ! See optimx manual to fit your needs. 
# In this case you may want to limit the natural frequency of the oscillator below 
# phenotype sampling rate and multiple start to avoid local optimum
fits<-optimx(params,fn = objfun,method=c("nlminb"))
fits
```

## Visualize fit

Let see the fit

```{r}
out<-SWDMrFit(model,fits)
```


```{r,warning=F}
# plot fitted lines
gg<-ggplot(aes(x=time,y=fit),data=cbind.data.frame(time=out$time,fit=out$y1))
gg<-gg+geom_line()
# Add expression points
gg <- gg + annotate("point",x=RNAExpression$Time,y=RNAExpression$Arntl)
gg<-gg +scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Arntl")

# Compute Sleep-wake driven value
SWdv<-SWdrivenValue(model,fits,c(0,24))
gg<-gg+annotate("text",  x=Inf, y = Inf, label = paste("SW driven:",round(SWdv,2)), vjust=1, hjust=1)
gg
```
Let's see the sinusoidal response, sleep-wake response and transient response


```{r}
out_response<-SWDMrFit(model,fits,method="Solve")
totalreponse<-out_response$circ_sol + out_response$trans_sol + out_response$SW_response + model@intercept

data_response=cbind.data.frame(time=out_response$time[-1],
                      response=c(out_response$circ_sol,fits$AmpSin * sin(2*pi/24*out_response$time[-1]+fits$PhiSin),
                                 out_response$trans_sol + out_response$SW_response,
                                 model@SWdist$Sleep*fits$Sleep + model@SWdist$Wake*fits$Wake),
                      value=c(rep("Circadian response",length(out_response$time[-1])),
                              rep("Circadian force",length(out_response$time[-1])),
                              rep("Sleep-Wake response",length(out_response$time[-1])),
                              rep("Sleep-Wake force",length(out_response$time[-1]))))

gg_resp<-ggplot(aes(time,y=response,color=value),data=data_response)+geom_line(size=1)
gg_resp<-gg_resp+scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Arntl - Responses")
gg_resp <- gg_resp + scale_color_manual(values=c("Circadian response" = rgb(224/255,160/255, 0/255),
                                                 "Sleep-Wake response" = rgb(20 /255,61/255, 89/255),
                                                 "Circadian force" =rgb(224/255,160/255, 0/255,.5),
                                                 "Sleep-Wake force" = rgb(20 /255,61/255, 89/255,.5)))
gg_resp
```



## Compute some statistics


```{r}
residuals<-SWDMrStats(model,out,detailed = T)$residuals
fitted<-SWDMrStats(model,out,detailed = T)$fitted
stats<-SWDMrStats(model,out,detailed = T)$stats
```


- Residual Sum of Square: RSS
- Negative Log Likelihood: NLL
- Bayesian Information Criterion: BIC
- BIC of a linear model with an intercept only: BIC_flat
- Bayes Factor between our model and a flat model: BayesFactor
- Akaike Information Criterion: AIC
- Number of samples in model: n
- Number of free parameters: k
- Variance of resiudals: ErrorVariance
- Kendall's tau between data points and fit: KendalTau


```{r}
stats
```


*Here n=56 while dataset contains 62 points. Because fit was performed up to Time96 while last points reach Time 216 and 222. If these points needed to be integrated, then SWdf should go up to time 222*


```{r}
par(mfrow=c(1,2))
plot(fitted,residuals);abline(h=mean(residuals))
qqnorm(residuals);qqline(residuals)
```


## Other dynamics


This model can also be applied to different dynamics like Homer1

```{r,warning=F}
model<-initDDHOmodel(swdmr,VarExp = "Homer1")
MeanGeneExprInBaseline<-mean(RNAExpression$Homer1[RNAExpression$Time<=48])
model<-FixIntercept(model,MeanGeneExprInBaseline)
model<-AddForce(model,"Wake")
model<-AddForce(model,"Sleep")
model<-AddSinF(model,FixPer = 24)
model<-SetYinitMode(model,mode = "Intercept_0")
model<-ReplicateDrivingForce(model,c(0,24.0),40)
model<-SetParametersModel(model)
objfun<-SWDMrGetEvalFun(model)
params<-c(omega=2*pi/24,loggamma=-1,Wake=0,Sleep=0,AmpSin=0,PhiSin=pi)
fits<-optimx(params,fn = objfun,method=c("nlminb"))
out<-SWDMrFit(model,fits)
gg<-ggplot(aes(x=time,y=fit),data=cbind.data.frame(time=out$time,fit=out$y1))
gg<-gg+geom_line()
gg <- gg + annotate("point",x=RNAExpression$Time,y=RNAExpression$Homer1)
gg<-gg +scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Homer1")

# Compute Sleep-wake driven value
SWdv<-SWdrivenValue(model,fits,c(0,24))
gg<-gg+annotate("text",  x=Inf, y = Inf, label = paste("SW driven:",round(SWdv,2)), vjust=1, hjust=1)

gg

# Stats
SWDMrStats(model,out,detailed = T)$stats

```


```{r}
out_response<-SWDMrFit(model,fits,method="Solve")
totalreponse<-out_response$circ_sol + out_response$trans_sol + out_response$SW_response + model@intercept

data_response=cbind.data.frame(time=out_response$time[-1],
                      response=c(out_response$circ_sol,fits$AmpSin * sin(2*pi/24*out_response$time[-1]+fits$PhiSin),
                                 out_response$trans_sol + out_response$SW_response,
                                 model@SWdist$Sleep*fits$Sleep + model@SWdist$Wake*fits$Wake),
                      value=c(rep("Circadian response",length(out_response$time[-1])),
                              rep("Circadian force",length(out_response$time[-1])),
                              rep("Sleep-Wake response",length(out_response$time[-1])),
                              rep("Sleep-Wake force",length(out_response$time[-1]))))

gg_resp<-ggplot(aes(time,y=response,color=value),data=data_response)+geom_line(size=1)
gg_resp<-gg_resp+scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Homer1 - Responses")
gg_resp <- gg_resp + scale_color_manual(values=c("Circadian response" = rgb(224/255,160/255, 0/255),
                                                 "Sleep-Wake response" = rgb(20 /255,61/255, 89/255),
                                                 "Circadian force" =rgb(224/255,160/255, 0/255,.25),
                                                 "Sleep-Wake force" = rgb(20 /255,61/255, 89/255,.25)))
gg_resp
```


Acot11

```{r,warning=F}
model<-initDDHOmodel(swdmr,VarExp = "Acot11")
MeanGeneExprInBaseline<-mean(RNAExpression$Acot11[RNAExpression$Time<=48])
model<-FixIntercept(model,MeanGeneExprInBaseline)
model<-AddForce(model,"Wake")
model<-AddForce(model,"Sleep")
model<-AddSinF(model,FixPer = 24)
model<-SetYinitMode(model,mode = "Intercept_0")
model<-ReplicateDrivingForce(model,c(0,24.0),40)
model<-SetParametersModel(model)
objfun<-SWDMrGetEvalFun(model)
params<-c(omega=2*pi/24,loggamma=-1,Wake=0,Sleep=0,AmpSin=0,PhiSin=pi)
fits<-optimx(params,fn = objfun,method=c("nlminb"))
out<-SWDMrFit(model,fits)
gg<-ggplot(aes(x=time,y=fit),data=cbind.data.frame(time=out$time,fit=out$y1))
gg<-gg+geom_line()
gg <- gg + annotate("point",x=RNAExpression$Time,y=RNAExpression$Acot11)
gg<-gg +scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Acot11")

# Compute Sleep-wake driven value
SWdv<-SWdrivenValue(model,fits,c(0,24))
gg<-gg+annotate("text",  x=Inf, y = Inf, label = paste("SW driven:",round(SWdv,2)), vjust=1, hjust=1)

gg

# Stats
SWDMrStats(model,out,detailed = T)$stats
```


```{r}
out_response<-SWDMrFit(model,fits,method="Solve")
totalreponse<-out_response$circ_sol + out_response$trans_sol + out_response$SW_response + model@intercept
data_response=cbind.data.frame(time=out_response$time[-1],
                      response=c(out_response$circ_sol,fits$AmpSin * sin(2*pi/24*out_response$time[-1]+fits$PhiSin),
                                 out_response$trans_sol + out_response$SW_response,
                                 model@SWdist$Sleep*fits$Sleep + model@SWdist$Wake*fits$Wake),
                      value=c(rep("Circadian response",length(out_response$time[-1])),
                              rep("Circadian force",length(out_response$time[-1])),
                              rep("Sleep-Wake response",length(out_response$time[-1])),
                              rep("Sleep-Wake force",length(out_response$time[-1]))))

gg_resp<-ggplot(aes(time,y=response,color=value),data=data_response)+geom_line(size=1)
gg_resp<-gg_resp+scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw() + ggtitle("Acot11 - Responses")
gg_resp <- gg_resp + scale_color_manual(values=c("Circadian response" = rgb(224/255,160/255, 0/255),
                                                 "Sleep-Wake response" = rgb(20 /255,61/255, 89/255),
                                                 "Circadian force" =rgb(224/255,160/255, 0/255,.5),
                                                 "Sleep-Wake force" = rgb(20 /255,61/255, 89/255,.5)))
gg_resp
```


# Fit a process-S dynamic on phenotype

This is a simplified version of process-S like driven phenotypes, see [J.Yeung github](https://github.com/jakeyeung/SleepDepAnalysis)

First we load data for time spent awake or asleep

## Data:

```{r}
# Contain a data.frame of time spent asleep or awake
data(SWdf)
head(SWdf)
```

We then load phenotypes containing also time of sampling corresponding with the sleep-wake data.frame

```{r}
data(RNAExpression)
head(RNAExpression)
```

Let's model the dynamics of Homer1

## Create model

We create a swdmr object containing sleep-wake data and phenotypes

```{r,warning=F}
swdmr <- SWDMr(SWdist=SWdf, Gexp=RNAExpression)
swdmr
```

We then initiate a model for a process-S dynamics

```{r,warning=F}
modelS<-initProcessSmodel(swdmr,VarExp = "Homer1")
```

We replicate the baseline time (time0-time24) 10x to have a stable baseline dynamics

```{r,warning=F}
modelS<-ReplicateDrivingForce(modelS,c(0,24.0),10)
```


We fix the starting point as mean expression level

```{r}
modelS<-SetYinitMode(modelS,"Fixed",mean(RNAExpression$Homer1))
```


We want to optimize our objective function to minimize the Residual Sum of Square (RSS)

```{r}
modelS<-SetFittingValue(modelS,value = "RSS")
```

Our model as the following free parameters:

```{r}
modelS
```

*We could fix some of these parameters*

```{r}
# Here, asymptotic value for time spent awake is 10 for the phenotype of interest
anothermodel<-FixAsymptWake(modelS,10)
anothermodel
```

## Optimization

We create an objective function:

```{r}
objfun<-SWDMrGetEvalFun(modelS)
```

We use optimx to optimize the free parameter to minimize the RSS

```{r}
paramsS<-c(AsympWake=6,AsympSleep=4,TauWake=10,TauSleep=10)
# Many other algorithms and options exist ! See optimx manual to fit your needs
fitsS<-optimx(paramsS,fn = objfun,method=c("Nelder-Mead"))
fitsS
```

## Visualize fit

Let see the fit

```{r}
outS<-SWDMrFit(modelS,fitsS)
```


```{r,warning=F}
# plot fitted lines
gg<-ggplot(aes(x=time,y=fit),data=cbind.data.frame(time=outS$time,fit=outS$y1))
gg<-gg+geom_line()
# Add expression points
gg <- gg + annotate("point",x=RNAExpression$Time,y=RNAExpression$Homer1)
gg<-gg +scale_x_continuous(breaks=seq(-24,96,by=12),limits = c(-24,96))+ theme_bw()
gg
```

## Compute some statistics


```{r}
residuals<-SWDMrStats(modelS,outS,detailed = T)$residuals
fitted<-SWDMrStats(modelS,outS,detailed = T)$fitted
stats<-SWDMrStats(modelS,outS,detailed = T)$stats
```


- Residual Sum of Square: RSS
- Negative Log Likelihood: NLL
- Bayesian Information Criterion: BIC
- BIC of a linear model with an intercept only: BIC_flat
- Bayes Factor between our model and a flat model: BayesFactor
- Akaike Information Criterion: AIC
- Number of samples in model: n
- Number of free parameters: k
- Variance of resiudals: ErrorVariance
- Kendall's tau between data points and fit: KendalTau


```{r}
stats
```


*Here n=56 while dataset contains 62 points. Because fit was performed up to Time96 while last points reach Time 216 and 222. If these points needed to be integrated, then SWdf should go up to time 222*


```{r}
par(mfrow=c(1,2))
plot(fitted,residuals);abline(h=mean(residuals))
qqnorm(residuals);qqline(residuals)
```


```{r}
sessionInfo()
```

